Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-11-20 threshold/draft.Rnw
Removed some words

Finding highly-correlated pairs among a large set of vectors is an important
part of many applications. For instance, subsets of highly-correlated vectors
of gene expression values may be used in the discovery of networks of genes
relevant to particular biological processes\cite{genomics}.  Identification of
highly-correlated pairs may be used in feature selection algorithms for machine
learning applications\cite{ml1, ml2} and are also important to time series and
image processing applications\cite{timeseries, svd-similarity}.

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to methods called \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. base
pruning rules on the frequency of mutually co-ocurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods only approximate a set containing the most highly-correlated
pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by the SVD.

Consider a real-valued data matrix $A$ consisting of $m$ observations of $n$
column vectors.  Denote the columns of $A$ as $a_j\in\mathcal{R}^m$ for
$j=1,2,\ldots,n$, $A=[a_1, a_2, \ldots, a_n]\in\mathcal{R}^{m\times n}$, and
let $k=\mathrm{rank}(A)$.  Assume that the mean of each column $a_j$ is zero
and that each column has unit norm $\|a_j\| = 1$. Here and below, $\|\cdot\|$
denotes the Euclidean vector norm. Then the Pearson sample correlation matrix
$\mathrm{cor}(A)=A^TA$.  Note that under these assumptions the sample
correlation and covariance matrices are the same.  Section \ref{irlba}
illustrates how to relax the unit norm and zero mean assumptions in practice.

@@

Lemma \ref{distlemma} equates the problem of finding pairs of highly correlated
column vectors from the matrix $A$ with a problem of finding pairs of vectors
sufficiently close together. For example, correlation values larger than 0.99
may only be associated with column vectors from the matrix $A$ whose Euclidean
distance is less than $\sqrt{0.02}$.  Vectors farther apart can not meet the
correlation threshold.

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} - e_j^Tv_{k})^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$--that is, the
$[j,k]$ entry in the matrix $V$.)

@@

Even a 1-d projected interval distance threshold may prune many possible
non-adjacent vector pairs with respect to the ordering defined by $P$ as
illustrated by the example in Figure \ref{fig1}.  However, this observation is
unlikely to rule out many adjacent pairs of vectors in most problems.  For
instance, the maximum distance between adjacent points shown in the example in
Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning occurs.

@@

Algorithm \ref{prune} by construction guarantees that no pruned vector pair
exceeds the given correlation threshold; that is, the pruning can not produce
any false negatives.  The algorithm typically prunes the vast majority of pairs
of vectors below the correlation threshold with truncated singular value
decompositions of only relatively low rank $p$.  The choice of $p$ is a balance
between work in computing a truncated SVD and work required by the $\ell$
matrix vector products of order $n\times p$ in step 4.

@@

The augmented implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlbar}, Matlab\cite{irlbam}, and Python\cite{irlbap}.

>>>>

Finding highly-correlated pairs among a large set of vectors is an important
part of many applications. For instance, subsets of highly-correlated vectors
of gene expression values may be used in the discovery of networks of genes
relevant to particular biological processes\cite{genomics}.  Identification of
highly-correlated pairs may be used in feature selection algorithms for machine
learning applications\cite{ml1, ml2} and are important to time series and
image processing applications\cite{timeseries, svd-similarity}.

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-ocurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by the SVD.

Consider a real-valued data matrix $A$ consisting of $m$ observations of $n$
column vectors.  Denote the columns of $A$ as $a_j\in\mathcal{R}^m$ for
$j=1,2,\ldots,n$, $A=[a_1, a_2, \ldots, a_n]\in\mathcal{R}^{m\times n}$, and
let $k=\mathrm{rank}(A)$.  Assume that the mean of each column $a_j$ is zero
and that each column has unit norm $\|a_j\| = 1$. Here and below, $\|\cdot\|$
denotes the Euclidean vector norm. Then the Pearson sample correlation matrix
$\mathrm{cor}(A)=A^TA$.  Note that under these assumptions the sample
correlation and covariance matrices are the same up to a constant multiple.
Section \ref{irlba} illustrates how to relax the unit norm and zero mean
assumptions in practice.

@@

Lemma \ref{distlemma} equates the problem of finding pairs of highly correlated
column vectors from the matrix $A$ with a problem of finding pairs of vectors
sufficiently close together. For example, correlation values larger than 0.99
may only be associated with column vectors from the matrix $A$ whose Euclidean
distance is less than $\sqrt{0.02}$.

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$
be any two columns of the matrix
$A$, $1\le i,j\le n$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} - e_j^Tv_{k})^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$--that is, the
$[j,k]$ entry in the matrix $V$.)

@@

Even a 1-d projected interval distance threshold may prune many possible
non-adjacent vector pairs with respect to the ordering defined by $P$ as
illustrated by the example in Figure \ref{fig1}.  However, this observation is
unlikely to rule out many adjacent pairs of vectors in most problems.  For
instance, the maximum distance between adjacent points shown in the example in
Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning of
adjacent pairs of vectors with respect to the ordering $P$ occurs.

@@

Algorithm \ref{prune} guarantees that no pruned vector pair exceeds the given
correlation threshold--the pruning does not produce false negatives.
The algorithm typically prunes the vast majority of pairs of vectors below the
correlation threshold with truncated singular value decompositions of only
relatively low rank $p$.  The choice of $p$ represents a balance between work in
computing a truncated SVD and work required by the $\ell$ matrix vector
products of order $n\times p$ in step 4.

@@

The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlbar}, Matlab\cite{irlbam}, and Python\cite{irlbap}.
